# An introduction to phylogenetic comparative methods in R
This tutorial is intended to get you familiar with the R environment for conducting
phylogenetic comparative analyses.
## I. Preparing the R environment

### Installing packages
To install a package directly from CRAN, use:
``` {r }
# install.packages("geiger")
```

Load the package using 'library'
``` {r }
library(geiger)
```

These days, many packages may not be on CRAN, but instead hosted on github, bitbucket etc.
To install these packages, it's useful to use the `devtools` package:
``` {r }
library(devtools)
install_github("richfitz/sowsear")
install_github("arborworkflows/aRbor")
library(sowsear)
```

For phylogenetic comparative methods, there are a number of packages available. One place
to find them all is the 'Phylogenetics Task View' run by Brian O'meara: (http://cran.r-project.org/web/views/Phylogenetics.html)
All of these packages can be installed at once using the following lines (however, please note that
this process can take a considerable amount of time)
``` {r }
# install.packages("ctv")
# install.views("Phylogenetics")
```

### Directory management
R looks for files in the 'working directory'. 
``` {r }
getwd()
dir.create("~/repos/SSBphylogenetics")
dir.create("~/repos/SSBphylogenetics/R")
dir.create("~/repos/SSBphylogenetics/data")
dir.create("~/repos/SSBphylogenetics/output")
setwd("~/repos/SSBphylogenetics/R/")
```

Save this script into the folder './R' and place your data files into the directory './data'
Now we are ready to read in data and tree files.

### II. Reading in a tree & tree data structures
There are a number of functions to read in phylogenetic tree data into R. We are going
to use as our example tree phylogeny of Muroid rodents (Schenk, Rowe & Steppan, 2013; Systematic Biology).
The tree can be downloaded from my github site, or it be accessed directly from treebase (submission 12303).
``` {r }
tree <- read.tree("../data/schenk.tre")
tree
```
We can plot the tree:
``` {r }
plot(tree)
plot(tree, type="fan", cex=0.25)
```
You may notice that the help file for `plot` is not very helpful for options for phylogenies: 
``` {r }
?plot
```
This is because `plot` is a "method" that behaves differently on objects of different classes. 
``` {r }
plot
methods(plot)
class(tree)
?plot.phylo
```

How is a tree stored? How can it be manipulated and modified? 
`str` is a useful commands for determining what is in an R object:
``` {r }
str(tree)
```
Each element of the list can be accessed by multiple means:
``` {r }
tree['tip.label']
tree[['tip.label']]
tree[[3]]
tree$tip.label
```
The structure of the tree is stored in the edge matrix:
``` {r }
tree$edge
```
And branch lengths are stored in the list element `edge.length`
``` {r }
tree$edge.length
```
Tips can be dropped with the function `drop.tip`:
``` {r }
#tree before:
tree
#tree after dropping two species:
drop.tip(tree, c("Eliomys_quercinus", "Sicista_tianshanica"))
```

### III. Matching a tree with a dataset
``` {r }
pantheria <- read.table("http://www.esapubs.org/archive/ecol/E090/184/PanTHERIA_1-0_WR05_Aug2008.txt",header=TRUE,sep="\t")
rownames(pantheria) <- gsub(" ", "_", pantheria$MSW05_Binomial)
head(pantheria)

td <- treedata(tree, pantheria)
td$data <- as.data.frame(td$data)
```

Experimental...Try the make.treedata function from the aRbor package
``` {r }
install_github("arborworkflows/aRbor")
require(aRbor)
td <- make.treedata(tree, pantheria)
td$dat[td$dat==-999] <- NA
td <- mutate(td, lnMass = log(X5.1_AdultBodyMass_g), lnBMR= log(X5.2_BasalMetRateMass_g), desert=ifelse(X28.1_Precip_Mean_mm < 21, 1, 0))
td <- select(td, lnMass, lnBMR, desert)
td <- filter(td, !is.na(lnMass), !is.na(lnBMR))

tree <- td$phy
dat <- td$dat
```

## IV. Visualization
Ape plotting:
``` {r }
plot(tree)
plot(tree, type="fan", cex=0.5)
plot(tree, show.tip.label=FALSE)
tiplabels(pch=21, bg=factor(dat$desert))
nodelabels(cex=0.75,bg="white" )
```

Plot a traitgram to visualize a continuous trait:
``` {r }
require(phytools)
phenogram(tree, setNames(dat$lnMass, tree$tip.label), spread.labels=FALSE)
```

Another useful tool is to plot the contrasts vs. the node height of the tree:
``` {r }
picMass <- pic(dat$lnMass, tree)
plot(tree, cex=0.5)
nodelabels(pch=21, bg=topo.colors(100)[round((picMass-min(picMass))/diff(range(picMass))*100,0)])
```

## V. Simple analyses of traits
### Discrete trait models
We generally use "continuous-time Markov models" to model discrete traits on the phylogeny. These models are
used to model transitions between discrete character states, whether they are the presence/absence of phenotypic trait
or nucleotide a specific site. We can fit such a model to a trait and a phylogeny using the `fitDiscrete` function in
in `geiger`. 
``` {r }
tdDiscrete <- filter(td, !is.na(desert))
trait <- setNames(tdDiscrete$dat$desert, tdDiscrete$phy$tip.label)+1
mER <- fitDiscrete(tdDiscrete$phy, trait, model="ER")
mARD <- fitDiscrete(tdDiscrete$phy, trait, model="ARD")
mER
mARD
```

### Continuous trait models
For continuous traits, we generally use "Gaussian models" that result in multivariate normal distributions.
The most basic and familiar of these is Brownian motion. We can fit a Brownian motion model using the function 
`fitContinuous` from the `geiger` package. 
``` {r }
trait <- setNames(dat$lnMass, tree$tip.label)
mBM <- fitContinuous(tree, trait, model="BM")
mBM
```
An more general model than the Brownian motion model is the Ornstein-Uhlenbeck model, which has one additional 
parameter describing the central tendency of traits to evolve toward an intermediate optimum (think of a rubber 
band pulling back towards a stick).
``` {r }
mOU <- fitContinuous(tree, trait, model="OU")
mBM
```
Another model is one in which the rate of evolution is initially rapid, but slows over time. This is called
the early-burst model (EB). Let's fit all models simultaneously using first 1) a loop and then 2) the lapply function.
This time, we'll use the BMR data instead of the mass data:
``` {r }
trait <- setNames(dat$lnBMR, tree$tip.label)
models <- c("BM", "OU", "EB")
mFits <- list()
for(i in 1:length(models)){
  mFits[[i]] <- fitContinuous(tree, trait, model=models[i])
}

#Above can also be accomplished by using lapply:
mFits <- lapply(1:length(models), function(i) fitContinuous(tree, trait, model=models[i]))
```

Model selection can be conducted by comparing AIC values:
``` {r }
aics <- lapply(mFits, function(x) x$opt$aic)
names(aics) <- models
aics
```

### Phylogenetic signal
Phylogenetic signal is covariation in traits between related species in the phylogeny.
There are several methods for testing for phylogenetic signal, remember to be cautious drawing
strong conclusions about biological processes from a simple summary statistic such as phylogenetic signal.
However, in general you can think of two sides of the continuum-- Strong phylogenetic signal indicating that 
closely related species are more similar to each other than other species. Whereas weak phylogenetic signal 
indicating that the variation among species is essentially random with respect to phylogeny.
#### Pagel's Lambda test
Pagel's lambda test compares the fit of a BM model on the original tree to a BM model with a lambda rescaled
tree. A lambda equal to 0 produces the following tree:
``` {r }
l0tree <- rescale(tree, "lambda", 0)
plot(l0tree)
```

A lambda equal to 1 is the original tree. We can estimate the best fitting tree as well.
``` {r }
trait <- setNames(dat$lnMass, tree$tip.label)
m1 <- fitContinuous(l0tree, trait, model = "BM")
m2 <- fitContinuous(tree, trait, model = "lambda")
lnlValues <- c(m1$opt$lnL, m2$opt$lnL)
names(lnlValues) <- c("Lambda fixed at zero", "Lambda estimated")
lambdaValue <- m2$opt$lambda
chisqTestStat <- 2 * (m2$opt$lnL - m1$opt$lnL)
chisqPVal <- pchisq(chisqTestStat, 1, lower.tail = F)
aiccScores <- c(m1$opt$aicc, m2$opt$aicc)
names(aiccScores) <- c("Lambda fixed at zero", "Lambda estimated")
res <- list(lnlValues = lnlValues, chisqTestStat = chisqTestStat, 
            chisqPVal = chisqPVal, aiccScores = aiccScores, lambdaValue = lambdaValue)
print(res)
```

Make this into a function:
``` {r }
physigLambda <- function(tree, trait){
  m1 <- fitContinuous(l0tree, trait, model = "BM")
  m2 <- fitContinuous(tree, trait, model = "lambda")
  lnlValues <- c(m1$opt$lnL, m2$opt$lnL)
  names(lnlValues) <- c("Lambda fixed at zero", "Lambda estimated")
  lambdaValue <- m2$opt$lambda
  chisqTestStat <- 2 * (m2$opt$lnL - m1$opt$lnL)
  chisqPVal <- pchisq(chisqTestStat, 1, lower.tail = F)
  aiccScores <- c(m1$opt$aicc, m2$opt$aicc)
  names(aiccScores) <- c("Lambda fixed at zero", "Lambda estimated")
  res <- list(lnlValues = lnlValues, chisqTestStat = chisqTestStat, 
              chisqPVal = chisqPVal, aiccScores = aiccScores, lambdaValue = lambdaValue)
  return(res)
}

physigLambda(tree, setNames(dat$lnBMR, tree$tip.label))
```

### Phylogenetic half-life
Phylogenetic half-life is a measure of phylogenetic signal proposed by Hansen et al. (2008) which
relies on an Ornstein-Uhlenbeck model of evolution. We will talk more about OU models later for studying 
adaptive evolution. For interpreting in terms of phylogenetic signal, it helps to consider 
in terms of the covariance between traits expected under a given phylogeny. For Brownian motion
the expected covariance is simply equal to the branch length of the tree. 
``` {r }
{
vcvBM <- vcvPhylo(tree, anc.nodes=FALSE)
dend <- as.dendrogram(as.hclust(tree))
heatmap(vcvBM, Rowv = dend, Colv = dend)

par(mfrow=c(3,3), mar=c(0,0,0,0))
image(vcvBM, xaxt="n", yaxt="n")
halflifeseq <- seq(50, 1, length.out=7)
for(i in 1:7){
  vcv <- vcvPhylo(rescale(tree, model="OU", log(2)/halflifeseq[i]), anc.nodes=FALSE)
  image(vcv, xaxt="n", yaxt="n")
}
iidData <- sapply(1:200, function(x) rnorm(length(tree$tip.label), mean(dat$lnMass), 2))
vcv <- var(iidData)
image(vcv, xaxt="n", yaxt="n")
}

mOU <- fitContinuous(tree, setNames(dat$lnMass, tree$tip.label), model="OU")
str(mOU)
halflife <- log(2)/mOU$opt$alpha
halflife/max(branching.times(tree))
```

#### PGLS - Determining relationship between traits
Phylogenetic generalized least squares is a method for assessing whether phenotypic traits are related to one 
another. This is a general method of the original independent contrasts methods proposed by Felsenstein (1985).
``` {r }
corBM <- corBrownian(1, phy=tree)
corOU <- corMartins(0.001, phy=tree)

pglsBM <- gls(lnBMR~lnMass, correlation=corBM, data=dat)
pglsOU <- gls(lnBMR~lnMass, correlation=corOU, data=dat)

summary(pglsBM)
summary(pglsOU)
```

## Exercises:
Fix the following errors: 


